Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-06-20 manuscript.Rnw
Update RNW

<<get_result_func>>=
# simple function to get rounded results with less typing
library(dplyr) # use to filter df prior to get_result in some cases
get_result <- function (results_table, col_name, row_name=1, signif=FALSE, digits=3) {
  if (signif == FALSE) {
    value <- round(results_table[row_name, col_name], digits=3)
  } else if (signif == TRUE) {
    value <- signif(results_table[row_name, col_name], digits=digits)
  }
  return(value)
}

@@

There were significant interaction effects between growth habit and elevation on all microclimatic variables except for minimum temperature (\tab \ref{supp_table:climate_model}). Slope and intercept for variables related to temperature were very similar between epiphytic and terrestrial dataloggers, but differed more strongly for those related to humidity (\tab \ref{supp_table:climate_model}, \fig \ref{fig:climate}). Minimum RH increased \Sexpr{get_result(slopes.results, "slope.epi", "min_RH", TRUE, 2)*100}\% per \SI{100}{\meter} elevation for epiphytic dataloggers, but was nearly flat (\Sexpr{get_result(slopes.results, "slope.ter", "min_RH", TRUE,2)*100}\% per \SI{100}{\meter}) for terrestrial dataloggers (linear model, $\rval^2 = \Sexpr{get_result(slopes.results, "r.squared", "min_RH")}$, $\pval < 0.001$). The Mt. Rotui \SI{830}{\meter} plot was a clear outlier, with greater mean temperature and lower minimum RH relative to other plots at similar elevation (outliers in  \fig \ref{fig:climate} at \circa \SI{800}{\meter}). Mt. Rotui is a solitary peak located between Cook's Bay and Oponohu Bay. It is more isolated with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \ref{fig:PCA}C, D). The first two PC axes explained \Sexpr{get_result(PCA.results, "Std_PCA_Dim.2", "cumulative_variance", digits=3, signif=TRUE)}\% and \Sexpr{get_result(PCA.results, "Phy_PCA_Dim.2", "cumulative_variance", digits=3, signif=TRUE)}\% of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In both analyses, stipe length and frond width were loaded on PC1, and SLA, degree of frond dissection, and frond length were loaded on PC2 (\fig \ref{fig:PCA}A, B, \tab \ref{supp_table:PCA}). Results for number of pinna pairs differed between the two analyses, tending toward PC2 in the standard PCA and PC1 in the phylogenetic PCA (\tab \ref{supp_table:PCA}).

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tab \ref{table:phylosig}, \fig \ref{fig:traits_tree}), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \ref{table:phylosig}a). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \ref{table:phylosig}a). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "hairs", digits=2, signif=TRUE)}$) or more conserved (gemmmae $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "gemmae", digits=2, signif=TRUE)}$; glands, $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "glands", digits=2, signif=TRUE)}$; morphotype, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "cordate_morph", digits=2, signif=TRUE)}$) than expected under a scenario of evolution by BM (\tab \ref{table:phylosig}b). Phylogenetic generalized linear mixed models (PGLMMs) also indicated significant signal in all gametophyte traits (\tab \ref{supp_table:PGLMM_gameto}). Growth habit showed phylogenetic signal nearly identical to that expected under BM ($\D = \Sexpr{get_result(binary_phylosig.results, "D", "epiphytic", digits=2, signif=TRUE)}$; \tab \ref{table:phylosig}b).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \ref{table:PIC_sporo}). Epiphytic species tend to have smaller values for all traits (\tab \ref{table:PIC_sporo}; contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, $\pval = \Sexpr{get_result(PIC.results, "pval", "stipe", digits=2, signif=TRUE)}$), as well as frond width ($\pval = \Sexpr{get_result(PIC.results, "pval", "width", digits=2, signif=TRUE)}$) and degree of laminae dissection ($\pval = \Sexpr{get_result(PIC.results, "pval", "dissection", digits=2, signif=TRUE)}$). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, $\pval = \Sexpr{get_result(pagel.results, "pval", "morph_binary", digits=2, signif=TRUE)}$; \tab \ref{table:pagel_gameto}). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (\tabs \ref{supp_table:PGLMM_sporo}, \ref{supp_table:PGLMM_gameto}).

@@

We observed a total of \Sexpr{ncol(comm)} species across all plots, including \Sexpr{length(traits$habit[traits$habit == "epiphytic"])} epiphytic and \Sexpr{length(traits$habit[traits$habit == "terrestrial"])} terrestrial species. Mean species richness was $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "mean", signif=TRUE)} \pm \stdev \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "sd", signif=TRUE)}$ \si{\species} per plot for epiphytes and $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "mean", signif=TRUE)} \pm \stdev \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "sd", signif=TRUE)}$ \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "pval", signif=TRUE)}$; \fig \ref{fig:FDIV}A). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \ref{fig:FDIV}B). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FRic", habit=="terrestrial"), "pval", signif=TRUE)}$), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FDiv", habit=="terrestrial"), "pval", signif=TRUE)}$), and functional evenness (FEve) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \ref{fig:FDIV}A). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain", signif=TRUE)}$, $\pval = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "p", signif=TRUE, digits=1)}$) and plots with greater minimum RH (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "p")}$). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mntd.obs.z"), "pval", signif=TRUE)}$) but not MPD ($\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mpd.obs.z"), "pval", signif=TRUE, digits=2)}$; \fig \ref{fig:FDIV}B). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \ref{fig:FDIV}A).

@@

% Table S5 
\begin{table}[ht]
\caption{Trait values and growth habit of ferns from Moorea, French Polynesia. For continuous trait values, means $\pm$ \stdev shown. For growth habit, E = epiphytic, T = terrestrial. For frond dissection, 1 = simple, 2 = pinnatifid or pinnatisect, 3 = 1-pinnate, 4 = 1- pinnate-pinnatifid, 5 = 2-pinnate, 6 = 2-pinnate-pinnatifid, 7 = 3- pinnate, 8 = 3-pinnate-pinnatifid, 9 = \textgreater 3-pinnate-pinnatifid, 10 = binpinnatifid or tripinnatifid.}
\label{supp_table:traitvals}
\end{table}

>>>>

<<get_result_func>>=
# simple function to get rounded results with less typing
library(dplyr) # use to filter df prior to get_result in some cases
get_result <- function (results_table, col_name, row_name=1, signif=FALSE, digits=2) {
  if (signif == FALSE) {
    value <- round(results_table[row_name, col_name], digits=digits)
  } else if (signif == TRUE) {
    value <- signif(results_table[row_name, col_name], digits=digits)
  }
  return(value)
}

@@

There were significant interaction effects between growth habit and elevation on all microclimatic variables except for minimum temperature (\tab \ref{supp_table:climate_model}). Slope and intercept for variables related to temperature were very similar between epiphytic and terrestrial dataloggers, but differed more strongly for those related to humidity (\tab \ref{supp_table:climate_model}, \fig \ref{fig:climate}). Minimum RH increased \Sexpr{get_result(slopes.results, "slope.epi", "min_RH", TRUE)*100}\% per \SI{100}{\meter} elevation for epiphytic dataloggers, but was nearly flat (\Sexpr{get_result(slopes.results, "slope.ter", "min_RH", TRUE)*100}\% per \SI{100}{\meter}) for terrestrial dataloggers (linear model, $\rval^2 = \Sexpr{get_result(slopes.results, "r.squared", "min_RH")}$, $\pval = \Sexpr{get_result(slopes.results, "pval", "min_RH", TRUE)}$). The Mt. Rotui \SI{830}{\meter} plot was a clear outlier, with greater mean temperature and lower minimum RH relative to other plots at similar elevation (outliers in  \fig \ref{fig:climate} at \circa \SI{800}{\meter}). Mt. Rotui is a solitary peak located between Cook's Bay and Oponohu Bay. It is more isolated with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \ref{fig:PCA}C, D). The first two PC axes explained \Sexpr{get_result(PCA.results, "Std_PCA_Dim.2", "cumulative_variance")}\% and \Sexpr{get_result(PCA.results, "Phy_PCA_Dim.2", "cumulative_variance")}\% of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In both analyses, stipe length and frond width were loaded on PC1, and SLA, degree of frond dissection, and frond length were loaded on PC2 (\fig \ref{fig:PCA}A, B, \tab \ref{supp_table:PCA}). Results for number of pinna pairs differed between the two analyses, tending toward PC2 in the standard PCA and PC1 in the phylogenetic PCA (\tab \ref{supp_table:PCA}).

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tab \ref{table:phylosig}, \fig \ref{fig:traits_tree}), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \ref{table:phylosig}a). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \ref{table:phylosig}a). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "hairs")}$) or more conserved (gemmmae $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "gemmae")}$; glands, $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "glands")}$; morphotype, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "cordate_morph")}$) than expected under a scenario of evolution by BM (\tab \ref{table:phylosig}b). Phylogenetic generalized linear mixed models (PGLMMs) also indicated significant signal in all gametophyte traits (\tab \ref{supp_table:PGLMM_gameto}). Growth habit showed phylogenetic signal nearly identical to that expected under BM ($\D = \Sexpr{get_result(binary_phylosig.results, "D", "epiphytic")}$; \tab \ref{table:phylosig}b).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \ref{table:PIC_sporo}). Epiphytic species tend to have smaller values for all traits (\tab \ref{table:PIC_sporo}; contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, $\pval = \Sexpr{get_result(PIC.results, "pval", "stipe", TRUE)}$), as well as frond width ($\pval = \Sexpr{get_result(PIC.results, "pval", "width", TRUE)}$) and degree of laminae dissection ($\pval = \Sexpr{get_result(PIC.results, "pval", "dissection", TRUE)}$). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, $\pval = \Sexpr{get_result(pagel.results, "pval", "morph_binary", TRUE)}$; \tab \ref{table:pagel_gameto}). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (\tabs \ref{supp_table:PGLMM_sporo}, \ref{supp_table:PGLMM_gameto}).

@@

We observed a total of \Sexpr{ncol(comm)} species across all plots, including \Sexpr{length(traits$habit[traits$habit == "epiphytic"])} epiphytic and \Sexpr{length(traits$habit[traits$habit == "terrestrial"])} terrestrial species. Mean species richness was $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "mean", signif=TRUE)} \pm \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "sd", signif=TRUE)}$ (\stdev) \si{\species}per plot for epiphytes and $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "mean", signif=TRUE)} \pm \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "sd", signif=TRUE)}$ (\stdev) \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "pval", signif=TRUE)}$; \fig \ref{fig:FDIV}A). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \ref{fig:FDIV}B). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FRic", habit=="terrestrial"), "pval", signif=TRUE)}$), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FDiv", habit=="terrestrial"), "pval", signif=TRUE)}$), and functional evenness (FEve) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \ref{fig:FDIV}A). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "p", signif=TRUE)}$) and plots with greater minimum RH (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "p")}$). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mntd.obs.z"), "pval", signif=TRUE)}$) but not MPD ($\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mpd.obs.z"), "pval", signif=TRUE)}$; \fig \ref{fig:FDIV}B). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \ref{fig:FDIV}A).

@@

% Table S5 
\begin{table}[ht]
\caption{Trait values and growth habit of ferns from Moorea, French Polynesia.}
\label{supp_table:traitvals}
\end{table}
